---
title: "Introduction to DataFrame"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to DataFrame}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

[![Travis-CI Build Status](https://travis-ci.org/wahani/dat.svg?branch=master)](https://travis-ci.org/wahani/dat)
[![codecov.io](https://codecov.io/github/wahani/dat/coverage.svg?branch=master)](https://codecov.io/github/wahani/dat?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/dat)](http://cran.r-project.org/package=dat)
[![Downloads](http://cranlogs.r-pkg.org/badges/dat?color=brightgreen)](http://www.r-pkg.org/pkg/dat)

```{r, results='asis', echo=FALSE}
cat(gsub("\\n   ", "", packageDescription("dat", fields = "Description")))
```

## Installation

```{r eval=FALSE}
devtools::install_github("wahani/dat")
```

## Why should you care?

This package links to dplyr functions and does not cover everything in it, so
why should you care:

- You probably have to rewrite all your dplyr / data.table code once you put it 
inside a package. I.e. working around non standard evaluation or find another
way to apiece R CMD check. And you don't like that.
- We work together and you have to understand code I wrote.

I cannot think of other reasons why you should care, of course if you just like
the syntax that's fine too.


## Examples:

I took the examples from the introductory vignette of dplyr. Things that are not supported:

- no rename
- no distinct
- no transmute

But you still work with data frames. So you can go back or mix in dplyr features
when you need them.

```{r}
library(nycflights13)
library(dplyr)
library(dat)

dat <- do.call(DataFrame, flights)
str(dat)
```

This function I use to compare if my code produces roughly (not the same class)
the same as dplyr code.

```{r}
myIdentical <- function(a, b) {
  l <- lapply(list(a, b), as.data.frame)
  do.call(identical, l)
}
```

### Filter rows

```{r}
dat %>%
  mutar(~1:10)

myIdentical(
  filter(flights, month == 1, day == 1),
  dat[~ month == 1 & day == 1]
)

myIdentical(
  slice(flights, 1:10),
  dat[~1:10]
)

myIdentical(
  # It is truly amazing how many times I tried to subset a data frame with
  # dat[1:10] and meant rows. Thats why this is working:
  dat[~1:10],
  dat[1:10, ]
)
```

### Sorting

```{r}
dat %>%
  mutar(~order(year, month, day))

myIdentical(
  arrange(flights, year, month, day),
  dat[~order(year, month, day)]
)
```


### Select cols

You can use characters and logicals to select cols of a *DataFrame*. Using numeric values is not supported; it is error prone and I have spent too many hours of my life debugging code where I relied on positions in a data frame.

```{r}
dat %>%
  mutar(c("year", "month", "day")) %>%
  mutar("year:day")

myIdentical(
  select(flights, year, month, day),
  dat[c("year", "month", "day")]
)

myIdentical(
  select(flights, year:day),
  dat["year:day"] # characters are passed into dplyr::select_
)
```

You can also pass in a function which checks if you want to select a column, e.g.
select all numeric cols:

```{r}
dat[is.numeric] # or mutar(dat, is.numeric)
```

Or select all cols with missing values in them:

```{r}
dat[function(x) any(is.na(x))]
```

### Mutate

```{r}
dat %>%
  mutar(gain ~ arr_delay - dep_delay,
        speed ~ distance / air_time * 60) %>%
  mutar("^gain|speed$")

myIdentical(
  mutate(flights,
         gain = arr_delay - dep_delay,
         speed = distance / air_time * 60),
  dat[gain ~ arr_delay - dep_delay,
      speed ~ distance / air_time * 60]
)
```

### Summarise

```{r}
dat %>%
  mutar(delay ~ mean(dep_delay, na.rm = TRUE), by = "month")

myIdentical(
  group_by(flights, month) %>% 
    summarise(delay = mean(dep_delay, na.rm = TRUE)),
  dat[delay ~ mean(dep_delay, na.rm = TRUE), by = "month"]
)

# dplyr:
delay1 <- flights %>%
  group_by(tailnum) %>%
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>%
  filter(count > 20, dist < 2000)

# dat #1:
dat %>%
  mutar(count ~ n(),
        dist ~ mean(distance, na.rm = TRUE),
        delay ~ mean(arr_delay, na.rm = TRUE),
        by = "tailnum") %>%
  mutar(~count > 20 & dist < 2000)

# dat #2:
delay2 <- dat[
  count ~ n(),
  dist ~ mean(distance, na.rm = TRUE),
  delay ~ mean(arr_delay, na.rm = TRUE),
  by = "tailnum"
  ] %>%
  .[~count > 20 & dist < 2000]

myIdentical(delay1, delay2)
```

