---
title: "Tools for Data Manipulation"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tools for Data Manipulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

[![Travis-CI Build Status](https://travis-ci.org/wahani/dat.svg?branch=master)](https://travis-ci.org/wahani/dat)
[![codecov.io](https://codecov.io/github/wahani/dat/coverage.svg?branch=master)](https://codecov.io/github/wahani/dat?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/dat)](http://cran.r-project.org/package=dat)
[![Downloads](http://cranlogs.r-pkg.org/badges/dat?color=brightgreen)](http://www.r-pkg.org/pkg/dat)

```{r, results='asis', echo=FALSE}
cat(gsub("\\n   ", "", packageDescription("dat", fields = "Description")))
```

## Installation

```{r eval=FALSE}
devtools::install_github("wahani/dat")
```

## Why should you care?

- You probably have to rewrite all your dplyr / data.table code once you put it 
inside a package. I.e. working around non standard evaluation or find another
way to satisfy R CMD check. And you don't like that.
- dplyr is not respecting the class of the object it operates on; the class
attribute changes on-the-fly.
- Neither dplyr nor data.table is playing nice with S4, but you really, really
want a S4 *data.table* or *tbl_df*.
- You like currying as in rlist and purrr.
- We work together and you have to understand code I wrote.

## map

What we can do with map:

```{r}
library(dat)
map(1:3, ~ .^2) # lapply
flatmap(1:3, ~ .^2) # lapply + unlist
flatmap(1:3, numeric(1) : x ~ x^2) # lapply + unlist + type check
map(ML(1:4, 4:1), integer : f(x, y) ~ rep(x, y)) # mapply + check return type
map(list(1:4, 4:1), 2:3) # subsetting on lists
map(ML(1:3, 11:13), c) # zip
map(ML(1:3, 11:13), c) %>% 
  { map(do.call(ML, .), c) } # unzip
dat <- data.frame(x = 1, y = "")
map(dat, x ~ x + 1, is.numeric) # only operates on numeric cols
```


## extract

Sum all even numbers from 1 to 10:

```{r}
is.even <- function(x) (x %% 2) == 0
sum((1:10)[is.even(1:10)])
1:10 %>% extract(~ . %% 2 == 0) %>% sum
1:10 %>% extract(is.even) %>% sum
```

Find all factors of 15:

```{r}
factors <- function(a, b = 1:a) b[a %% b == 0]
factors(15)
extract(1:15, ~ 15 %% . == 0)
```

Find all relative prime numbers:

```{r}
gcd <- function(a, b) {
  .gcd <- function(a, b) if (b == 0) a else Recall(b, a %% b)
  map(ML(a, b), .gcd) %>% unlist
}

extract(1:10, x ~ gcd(x, 10) == 1)
```

Find real prime numbers:

```{r}
isPrime <- function(n) {
  .isPrime <- function(n) {
    iter <- function(i) {
      if (i * i > n) TRUE
      else if (n %% i == 0 || n %% (i + 2) == 0) FALSE
      else Recall(i + 6)
    }
    if (n <= 1) FALSE
    else if (n <= 3) TRUE
    else if (n %% 2 == 0 || n %% 3 == 0) FALSE
    else iter(5)
  }
  flatmap(n, logical(1) : x ~ .isPrime(x))
}

extract(1:10, isPrime)
```

## Data Frame

I took the examples from the introductory vignette of dplyr. Things that are not
supported:

- rename
- distinct
- transmute

But you still work with data frames. So you can go back or mix in dplyr features
when you need them.

```{r}
library(nycflights13)
library(dplyr)

dat <- do.call(DataFrame, flights)
str(dat)
```

### Filter

```{r results='hide'}
filter(flights, month == 1, day == 1) # dplyr
dat[~ month == 1 & day == 1]          # #1
mutar(dat, ~ month == 1 & day == 1)   # #2

slice(flights, 1:10)                  # dplyr
dat[~1:10]                            # #1
dat %>% mutar(~1:10)                  # #2
```

```{r}
dat[1:10, ]                           # standard things
```

### Sorting

```{r results='hide'}
arrange(flights, year, month, day)    # dplyr
dat[~order(year, month, day)]         # #1
```

```{r}
dat %>% mutar(~order(year, month, day)) # #2
```

### Select cols

You can use characters and logicals to select cols of a *DataFrame*. Using
numeric values is not supported; it is error prone and I have spent too many
hours of my life debugging code where I relied on positions in a data frame.

```{r results='hide'}
select(flights, year, month, day)       # dplyr 
select(flights, year:day)               # dplyr

# characters are passed into dplyr::select_:
dat %>%
  mutar(c("year", "month", "day")) %>%
  mutar("year:day")

dat[c("year", "month", "day")]["year:day"]
```

You can also pass in a function which checks if you want to select a column, e.g.
select all numeric cols:

```{r}
dat[is.numeric] # or mutar(dat, is.numeric)
```

Or select all cols with missing values in them:

```{r results='hide'}
dat[function(x) any(is.na(x))]
```

### Mutate

```{r results='hide'}
flights %>% 
  mutate(gain = arr_delay - dep_delay,
         speed = distance / air_time * 60)

dat[gain ~ arr_delay - dep_delay,
    speed ~ distance / air_time * 60]
```

```{r}
dat %>%
  mutar(gain ~ arr_delay - dep_delay,
        speed ~ distance / air_time * 60) %>%
  mutar("^gain|speed$") # I assume a regex if the character begins with '^'
```

### Summarise

```{r results='hide'}
group_by(flights, month) %>% 
    summarise(delay = mean(dep_delay, na.rm = TRUE))

dat[delay ~ mean(dep_delay, na.rm = TRUE), 
    by = "month"]
```

```{r}
mutar(dat, 
      delay ~ mean(dep_delay, na.rm = TRUE), 
      by = "month")
```

```{r}
# dplyr:
delay1 <- flights %>%
  group_by(tailnum) %>%
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>%
  filter(count > 20, dist < 2000)

# dat #1:
delay2 <- dat[
  count ~ n(),
  dist ~ mean(distance, na.rm = TRUE),
  delay ~ mean(arr_delay, na.rm = TRUE),
  by = "tailnum"
  ] %>%
  .[~count > 20 & dist < 2000]
```

```{r}
# dat #2:
dat %>%
  mutar(count ~ n(),
        dist ~ mean(distance, na.rm = TRUE),
        delay ~ mean(arr_delay, na.rm = TRUE),
        by = "tailnum") %>%
  mutar(~count > 20 & dist < 2000)
```

### Split-apply-combine

```{r}
# The naive split-apply-combine
map(dat, mutar, By("month"), count ~ n()) %>% 
  mutar(27003:27006, j = c("month", "count"))
```


### Mixing dplyr and mutar

```{r}
dat %>%
  group_by(tailnum) %>%
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>%
  mutar(~count > 20 & dist < 2000)
```

### Example with S4 and data.table

Let's define a S4 class inheriting from data.table and do something with it.

```{r}
library(data.table)
setOldClass(c("data.table", "data.frame"))

setClass("DataTable", "data.table")

DataTable <- function(...) {
  new("DataTable", data.table::data.table(...))
}

setMethod("[", "DataTable", mutar)

set.seed(1)
dat <- DataTable(id = rep(letters[1:2], each = 10), x = 101:120, y = rnorm(20))
dat[1:2, ]
dat %>% 
  mutar(~1:2, y ~ runif(2))

dat %>%
  mutar(sumOfX ~ sum(x), by = "id")

map(dat, mutar, By("id"), sumOfX ~ sum(x))[~1:2]
map(dat, dt ~ DataTable(sumOfX = sum(dt$x)), By("id"))
```

