---
title: "Tools for Data Manipulation"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tools for Data Manipulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

[![Travis-CI Build Status](https://travis-ci.org/wahani/dat.svg?branch=master)](https://travis-ci.org/wahani/dat)
[![codecov.io](https://codecov.io/github/wahani/dat/coverage.svg?branch=master)](https://codecov.io/github/wahani/dat?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/dat)](http://cran.r-project.org/package=dat)
[![Downloads](http://cranlogs.r-pkg.org/badges/dat?color=brightgreen)](http://www.r-pkg.org/pkg/dat)

```{r, results='asis', echo=FALSE}
cat(gsub("\\n   ", "", packageDescription("dat", fields = "Description")))
```

## Installation

```{r eval=FALSE}
devtools::install_github("wahani/dat")
```

## Why should you care?

- You probably have to rewrite all your dplyr / data.table code once you put it 
inside a package. I.e. working around non standard evaluation or find another
way to satisfy `R CMD check`. And you don't like that.
- `dplyr` is not respecting the class of the object it operates on; the class
attribute changes on-the-fly.
- Neither `dplyr` nor `data.table` is playing nice with S4, but you really,
really want a S4 *data.table* or *tbl_df*.
- You like currying as in `rlist` and `purrr`.


## A link to `dplyr`

The examples are from the introductory vignette of `dplyr`. Things that are not
supported:

- rename
- distinct
- transmute

But you still work with data frames: so you can simply mix in dplyr features 
whenever you need them. The functions `filtar`, `mutar` and `sumar` are `R CMD 
check` friendly replacements for the corresponding versions in `dplyr`. For 
`select` you can use `extract`. The function names are chosen so that they are 
similar but do not conflict with `dplyr`s function names - so `dplyr` can be
savely attached to the search path.

```{r}
library(nycflights13)
library(dat)
```

### Select rows

`filtar` can be used as a replacement for `filter` and `slice`. When you
reference a variable in the data itself, you can indicate this by using a one
sided formula.

```{r results='hide'}
filtar(flights, ~ month == 1 & day == 1)  
filtar(flights, 1:10)
```

And for sorting:

```{r}
filtar(flights, ~order(year, month, day))
```


### Select cols

You can use characters, logicals, regular expressions and functions to select
columns. Regular expressions are indicated by a leading "^". Character are
simply passed to `dplyr::select_`.

```{r results='hide'}
flights %>%
  extract(c("year", "month", "day")) %>%
  extract("year:day") %>%
  extract("^day$") %>%
  extract(is.numeric)
```


### Mutate

```{r results='hide'}
mutar(
  flights,
  gain ~ arr_delay - dep_delay,
  speed ~ distance / air_time * 60
)
```

Grouping data is handled within `mutar`:

```{r}
mutar(flights, n ~ n(), by = "month") %>%
  extract("n") %>% 
  unique
```


### Summarise

```{r results='hide'}
sumar(flights, delay ~ mean(dep_delay, na.rm = TRUE), by = "month")
```


### Same operations on different columns

```{r}
sumar(
  flights,
  FL(.n ~ mean(.n, na.rm = TRUE), 
     .n = c("arr_delay", "dep_delay")), 
  by = "month"
)
```


## A link to S4

Using this package you can create S4 classes to contain a data frame (or a 
data.table) and use the interface to `dplyr`. Both `dplyr` and `data.table` do 
not support integration with S4. The main function here is `mutar` which is
generic enough to link to subsetting of rows and cols as well as mutate and 
summarise. In the background `dplyr`s ability to work on a `data.table` is being
used.

```{r eval = FALSE}
library(data.table)

setClass("DataTable", "data.table")

DataTable <- function(...) {
  new("DataTable", data.table::data.table(...))
}

setMethod("[", "DataTable", mutar)

dtflights <- do.call(DataTable, flights)

dtflights[1:10, "year:day"]
dtflights[n ~ n(), by = "month"] %>% extract("n") %>% unique
dtflights[n ~ n(), sby = "month"]

dtflights %>%
  filtar(~month > 6) %>%
  mutar(n ~ n(), by = "month") %>%
  sumar(n ~ dplyr::first(n), by = "month")
```


## Working with vectors

Inspired by `rlist` and `purrr` some low level operations on vectors are
supported. The aim here is to integrate syntactic sugar for anonymous functions.
Furthermore the functions should support the use of pipes.

- `map` and `flatmap` as replacements for the apply functions
- `extract` for subsetting
- `replace` for replacing elements in a vector

What we can do with map:

```{r eval=FALSE}
map(1:3, ~ .^2)
flatmap(1:3, ~ .^2)
map(1:3 ~ 11:13, c) # zip
dat <- data.frame(x = 1, y = "")
map(dat, x ~ x + 1, is.numeric)
```

What we can do with extract:

```{r eval=FALSE}
extract(1:10, ~ . %% 2 == 0) %>% sum
extract(1:15, ~ 15 %% . == 0)
l <- list(aList = list(x = 1), aAtomic = "hi")
extract(l, "^aL")
extract(l, is.atomic)
```

What we can do with replace:

```{r eval=FALSE}
replace(c(1, 2, NA), is.na, 0)
replace(c(1, 2, NA), rep(TRUE, 3), 0)
replace(c(1, 2, NA), 3, 0)
replace(list(x = 1, y = 2), "x", 0)
replace(list(x = 1, y = 2), "^x$", 0)
replace(list(x = 1, y = "a"), is.character, NULL)
```
